---
title: "Linear Regression and the Conditional Expectation Function (CEF)"
author: "Manoel Galdino"
date: "2024-01-26"
output: ioslides_presentation
---

# Introduction

- **Objective of the Lecture:** Define and explain the concept of the Conditional Expectation Function (CEF) and its relevance in political science research.
- **Brief Overview of CEF:** Introduction to the concept of expectation in statistics and how it extends to conditional expectation.
- **Importance in Political Science:** Discuss how CEF is used to understand relationships between variables in political science.

# Theoretical Background

## Basic Statistical Concepts
Review of expectation, variance, and their properties.

## Definition of CEF
Detailed explanation of CEF and its mathematical formulation.

## Interpreting CEF
Understanding what CEF tells us about the relationship between two variables.

# CEF in Regression Analysis

## Linear Regression Model
Overview of how linear regression is a special case of CEF.

## Assumptions of Linear Regression
Discuss assumptions and how they relate to CEF.

## Interpreting Coefficients
Understanding coefficients in regression as conditional expectations.

# Practical Applications in Political Science

## Case Studies
Presenting real-world examples where CEF is applied in political science research.

## Wage and Demographics

```{r import-pnad, echo=FALSE, warning = FALSE, message=FALSE, cache=TRUE}
# Carregue o pacote PNADcIBGE
# library(PNADcIBGE)
# Importe os dados desejados
# data <- get_pnadc(year=2017,
#                   quarter=4,
#                   selected=FALSE,
#                   vars=c("Ano", "Trimestre", "UF", "V2007", "VD4020", "VD4035", "V1028", "V2010", "V3009A"),
#                   design=FALSE, savedir=tempdir())
# 
library(tidyverse)
# data <- data %>%
#   select(Ano, Trimestre, UF, V2007, VD4020, VD4035, V1028, V2010, V3009A)
# # Renaming variables:
# data <- data %>%
#   rename(gender = V2007,
#          income = VD4020,
#          work_hours = VD4035,
#          race = V2010,
#          education = V3009A)
# saveRDS(data, file="pnad.rds")
data <- readRDS(file="pnad.rds")
```
```{r processing-pnad, echo=FALSE, message=FALSE, cache=TRUE, warning=FALSE}

library(ggplot2)

levels <- c("College elementary or lower", "middle school", "high school", "Post-grad")


# recode
data <- data %>%
  mutate(education1 = as.character(education),
         education1 = ifelse(grepl("(elementar|fundamental|lfabetiza|-escola)", education1), "elementary or lower",
                             ifelse(grepl("gin",education1), "middle School",
                                    ifelse(grepl("supletivo do 1", education1), "middle school",
                                           ifelse(grepl("Antigo cient|supletivo do 2|Regular do ensino m", education1), "high school", education1))))) %>%
  mutate(education1= ifelse(grepl("Superior", education1 ), "College",
                            ifelse(grepl("Mestrado|Especializa|Doutorado", education1 ), "Post-grad", education1)),
         education1 = factor(education1, levels = levels, ordered=TRUE))

df <- data %>%
  filter(!is.na(income)) %>%
  filter(!is.na(work_hours)) %>%
  filter(income > 0) %>%
  filter(work_hours > 0) %>%
  mutate(wage = income/(4.5*work_hours)) %>%
    mutate(log_wage = log(wage)) %>%
  mutate(gender = as.character(gender)) 
```
### Wage - density
```{r wage-pnad, echo=FALSE, results='asis', message=FALSE, cache=TRUE, warning=F, fig.cap = "Distribuição dos salários" }


p1 <- df %>%
  ggplot(aes(wage)) + geom_density(aes(weight=V1028)) + theme_bw(base_size = 22)

print(p1)

```
## Log Wage

```{r wage-pnad-log, echo=FALSE, results='asis', message=FALSE, cache=TRUE, fig.cap = "Distribuição do log dos salários" }

p2 <- df %>%
  ggplot(aes(log_wage)) + geom_density(aes(weight=V1028)) + theme_bw(base_size = 22)

print(p2)

```

## Wage and Gender


```{r wage-cond-expec, echo=FALSE, results='asis', message=FALSE, warning=F, cache=TRUE, fig.cap = "Distribuição dos salários" }


p3 <- df %>%
  ggplot(aes(x=log_wage)) +
  geom_density(aes(weight = V1028, group=gender, colour=gender)) + theme_bw(base_size = 22)

print(p3)

```

## Wage and Education
```{r wage-cond-expec-educ, results='asis', message=FALSE, warning=F, cache=TRUE, fig.cap = "Distribuição dos salários" }


p3 <- df %>%
  ggplot(aes(x=log_wage)) +
  geom_density(aes(weight = V1028, group=education1, colour=education1)) + theme_bw(base_size = 22)

print(p3)

```